---
title: "Lab 2"
subtitle: "Intro to linear regression"
author: "Juan Carlos Villaseñor-Derbez"
date: "23 de enero de 2018"
output:
  html_document:
    code_folding: show
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r}
suppressPackageStartupMessages({
  library(stargazer)
  library(ggfortify)
  library(tidyverse)
})
```


```{r}
SmallHHfile <- read.csv("../Lab1/SmallHHfile.csv", header=TRUE)
```

# Describe data

## Look at the structure of the file

```{r, results = "asis"}
skimr::skim(SmallHHfile)
```


## Summary statistics with `stargazer`

```{r}
stargazer(SmallHHfile, type = "text", title="Descriptive statistics", median=TRUE, digits=2, out="table1.txt")
```

## Fit a model

```{r, results = "asis"}
#  The model in slides 28 and 29 from lecture
HHPMT.lm = lm(TotDist ~ HHSIZ , data=SmallHHfile)
stargazer(HHPMT.lm, single.row = T, type = "html")
```

```{r}
anova(HHPMT.lm) # for slide 33 in lecture notes
```

# Simple comparison of means and sample size example (about slide 70 of intro to regression)

## `n = 10`

### Sample ten random values from `SmallHHFile`

```{r}
TEN <- SmallHHfile[sample(nrow(SmallHHfile), 10), ] # randomly draw a sample of ten rows from SmallHHfile
```

### Obtain descriptive parameters

```{r}
xbar10 <- mean(TEN$MilesPr)            # sample mean 
mu0 <- mean(SmallHHfile$MilesPr)             # hypothesized value 
sigma <- sd(SmallHHfile$MilesPr)            # population standard deviation 
n <- nrow(TEN)                 # sample size 

(z10 <-  (xbar10-mu0)/(sigma/sqrt(n)))
```

### compute critical values at 5% significance

```{r}
alpha = .05 # set the confidence level we want
z.half.alpha = qnorm(1 - alpha/2) # area under the normal curve 
c( -z.half.alpha, z.half.alpha)  #critical values
```

### Repeat with `n = 100`

```{r}
HUNDRED <- SmallHHfile[sample(nrow(SmallHHfile), 100), ] # randomly draw a sample of ten rows from SmallHHfile
```

### Obtain descriptive parameters

```{r}
xbar100 <- mean(HUNDRED$MilesPr)            # sample mean 
mu0 <- mean(SmallHHfile$MilesPr)             # hypothesized value 
sigma <- sd(SmallHHfile$MilesPr)            # population standard deviation 
n <- nrow(HUNDRED)                 # sample size 

(z100 <- (xbar100-mu0)/(sigma/sqrt(n)))
```

### compute critical values at 5% significance

```{r}
alpha = .05 # set the confidence level we want
z.half.alpha = qnorm(1 - alpha/2) # area under the normal curve 
c( -z.half.alpha, z.half.alpha)  #critical values
```

## When $\sigma$ is unknown

We use a t-distribution

```{r}

xbar10 <- mean(TEN$MilesPr)            # sample mean 
mu0 <- mean(SmallHHfile$MilesPr)             # hypothesized value 
s10 <- sd(TEN$MilesPr)            # population standard deviation 
n <- nrow(TEN) 

# test statistic for TEN
(t10 = (xbar10-mu0)/(s10/sqrt(n)))

```

### compute critical values at 5% significance

```{r}
# the above shows a z = -1.31147
alpha <-  .05 
t.half.alpha <-  qt(1 - alpha/2, df=n-1) 
c(- t.half.alpha, t.half.alpha) 
```

# Linear regression models in R (one explanatory variable)

## Inspect the model that we fit earlier

```{r}
HHPMT.fit = fitted(HHPMT.lm)
plot(SmallHHfile$HHSIZ, HHPMT.fit, type="b" )
```

## Inspect with data

```{r}
plot(TotDist ~ HHSIZ, data = SmallHHfile)
abline(HHPMT.lm)
```

```{r}
cool1 <- ggplot(SmallHHfile, aes(x=HHSIZ, y=TotDist)) + geom_point(shape=1) + geom_smooth(method=lm)
cool1 <- cool1 + scale_x_continuous(name = "Household Size") +
  scale_y_continuous(name = "Miles Travelled in a Day by each household")
cool1

# add a title to the graph
cool1 <- cool1+ ggtitle("Miles travelled Regression Line in CHTS 43431 households")

# use annotate to add a box with the regression coefficients and R square
cool1<- cool1 + annotate("rect", xmin = 7, xmax = 8, ymin = 4000, ymax = 6000, fill="white", colour="red") +
  annotate("text", x=7.5, y=4500, label = "R^2 == 0.06", parse=T) + annotate("text", x=7.5, y=5500, label = "alpha == 12.2", parse=T) +
  annotate("text", x=7.5, y=5000, label = "beta == 21.7", parse=T)
cool1


# The following will create an object called equation fro the linear regression lm object and also extract the r-square

equation = function(x) {
  lm_coef <- list(a = round(coef(x)[1], digits = 3),
                  b = round(coef(x)[2], digits = 3),
                  r2 = round(summary(x)$r.squared, digits = 2));
  lm_eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(R)^2~"="~r2,lm_coef)
  as.character(as.expression(lm_eq));                 
}

# then we insert in the cool graph the equation from the object lm

cool1 <- cool1 + annotate("text", x = 2, y = 5000, label = equation(HHPMT.lm), parse = TRUE)
cool1
```

```{r}
autoplot(HHPMT.lm, label.size = 3)
```










