---
title: "Assignment 2"
subtitle: "Matrix Linear Regression with R and Diagnostics"
author: "Juan Carlos Villaseñor-Derbez"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r, echo = F}
suppressPackageStartupMessages({
  library(matlib)
  library(tidyverse)
})
```


# Part 1

Estimate a linear regression model with matrix algebra. The model coefficients are given by: 

$$\hat{\beta} = (X'X)^{-1} (X'Y)$$

Here, $X$ is the matrix:

$$
X = \left[\begin{array}{cccc}
    1 & 2 & 43 & 1 \\
    1 & 3 & 42 & 1 \\
    1 & 1 & 43 & 1 \\
    1 & 5 & 54 & 1 \\
    1 & 9 & 61 & 0 \\
    1 & 11 & 35 & 0 \\
    1 & 11 & 52 & 0 \\
    1 & 11 & 86 & 0 \\
    1 & 12 & 45 & 0 \\
    1 & 12 & 44 & 0 \\
    1 & 12 & 34 & 0 \\
\end{array}\right]
$$


And $Y$ is given by:

$$
Y = \left[\begin{array}{c}
4 \\
7 \\
3 \\
9 \\
17 \\
27 \\
13 \\
121 \\
10 \\
11 \\
23
\end{array}\right]
$$

Define X and Y:

```{r}
x1 <- rep(1L, 11)
x2 <- c(2L, 3L, 1L, 5L, 9L, 11L, 11L, 11L, 12L, 12L, 12L)
x3 <- c(43L, 42L, 43L, 54L, 61L, 35L, 52L, 86L, 45L, 44L, 34L)
x4 <- c(1L, 1L, 1L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 0L)

X <- cbind(x1, x2, x3, x4)

Y <- c(4L, 7L, 3L, 9L, 17L, 27L, 13L, 121L, 10L, 11L, 23L)
```

Calculate $X'$:

```{r}
XT <- t(X)
```

$$
X' =
\left[\begin{array}{ccccccccccc}
1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
2 & 3 & 1 & 5 & 9 & 11 & 11 & 11 & 12 & 12 & 12 \\
43 & 42 & 43 & 54 & 61 & 35 & 52 & 86 & 45 & 44 & 34 \\
1 & 1 & 1 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0
\end{array}\right]
$$

Calculate $X'X$:

```{r}
XTX <- XT %*% X
```


$$
X'X = 
\left[\begin{array}{ccccccccccc}
1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
2 & 3 & 1 & 5 & 9 & 11 & 11 & 11 & 12 & 12 & 12 \\
43 & 42 & 43 & 54 & 61 & 35 & 52 & 86 & 45 & 44 & 34 \\
1 & 1 & 1 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0
\end{array}\right]
\times
\left[\begin{array}{cccc}
    1 & 2 & 43 & 1 \\
    1 & 3 & 42 & 1 \\
    1 & 1 & 43 & 1 \\
    1 & 5 & 54 & 1 \\
    1 & 9 & 61 & 0 \\
    1 & 11 & 35 & 0 \\
    1 & 11 & 52 & 0 \\
    1 & 11 & 86 & 0 \\
    1 & 12 & 45 & 0 \\
    1 & 12 & 44 & 0 \\
    1 & 12 & 34 & 0 \\
\end{array}\right]
$$

$$
X'X = \left[
  \begin{array}{cccc}
    11 & 89 & 539 & 4 \\
    89 & 915 & 4453 & 11 \\
    539 & 4453 & 28541 & 182 \\
    4 & 11 & 182 & 4
  \end{array}
\right]
$$

Calculate $(X'X)^{-1}$:

```{r}
XTXi <- inv(XTX)
```

$$
(X'X)^{-1} = \left[
  \begin{array}{cccc}
    11 & 89 & 539 & 4 \\
    89 & 915 & 4453 & 11 \\
    539 & 4453 & 28541 & 182 \\
    4 & 11 & 182 & 4
  \end{array}
\right]^{-1}
$$

$$
(X'X)^{-1} = \left[
  \begin{array}{cccc}
10.47 & -0.77 & -0.03 & -6.79 \\
-0.77 &  0.06 &  0 &  0.55 \\
-0.03 &  0 &  0 &  0 \\
-6.79 &  0.55 &  0 &  5.08 \\
  \end{array}
\right]
$$
\clearpage

Now we need $X'Y$

```{r}
XTY <- XT %*% Y
```

$$
X'Y = 
\left[\begin{array}{ccccccccccc}
1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 \\
2 & 3 & 1 & 5 & 9 & 11 & 11 & 11 & 12 & 12 & 12 \\
43 & 42 & 43 & 54 & 61 & 35 & 52 & 86 & 45 & 44 & 34 \\
1 & 1 & 1 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0
\end{array}\right]
\times
\left[\begin{array}{c}
4 \\
7 \\
3 \\
9 \\
17 \\
27 \\
13 \\
121 \\
10 \\
11 \\
23
\end{array}\right]
$$

$$
X'Y = \left[\begin{array}{c}
245 \\
2529 \\
15861 \\
23
\end{array}\right]
$$

Finally, multiply $(X'X)^{-1} \times (X'Y)$ to get $\hat\beta$:

```{r}
beta <- XTXi %*% XTY
```

$$
\begin{split}
\hat{\beta} &= (X'X)^{-1} \times (X'Y) \\
 &= \left[
  \begin{array}{cccc}
10.47 & -0.77 & -0.03 & -6.79 \\
-0.77 &  0.06 &  0 &  0.55 \\
-0.03 &  0 &  0 &  0 \\
-6.79 &  0.55 &  0 &  5.08 \\
  \end{array}
\right] \times
 \left[\begin{array}{c}
245 \\
2529 \\
15861 \\
23
\end{array}\right] \\
 & =  \left[\begin{array}{c}
-82.32 \\
2.31 \\
1.72 \\
2.98
\end{array}\right]
\end{split}
$$

\clearpage

We can compare these results to what we would have obtained using the `lm` function:

```{r, results = "asis"}
lm(formula = Y ~ X - 1) %>% 
  stargazer::stargazer(single.row = T,
                       header = F,
                       title = "Regression coefficients estimated with the lm function")
```

\clearpage

# Part 2

## Estimate model from part 3 in Assignment 1

\begin{multline}
HTRIPS = \beta_0 + \beta_2 HHSIZ + \beta_3 HHVEH + \beta_4 TrpPrs + \beta_5 INCOM \\ + \beta_6 Mon + \beta_7 Tue + \beta_8 Wed + \beta_9 Thu + \beta_{10} Fri + \beta_{11} Sat +\epsilon
\end{multline}

```{r}
SmallHHfile <- read_csv("../../Labs/Lab1/SmallHHfile.csv", col_types = cols())

model <- lm(formula = HTRIPS ~ HHSIZ + HHVEH + TrpPrs +INCOM
             + Mon + Tue + Wed + Thu + Fri + Sat , data = SmallHHfile)
```

## Regression table

```{r, results = "asis"}
stargazer::stargazer(model, single.row = T, header = F)
```

\clearpage

## Summary of the model

```{r}
summary(model)
```

## ANOVA

```{r}
anova(model)
```

## 	Overall	Significance	of	the	Multiple Regression	Model

$$H_0 = \beta_0 = \beta_1 = ... = \beta_i = 0$$

$$H_1 = \text{at least one} \beta_i \neq 0$$

```{r}
(F_wald <- lmtest::waldtest(model))
```

This indicates that at least one of the coefficients is different from zero (`r paste0("F(df = ", F_wald$Res.Df[2]-F_wald$Res.Df[1], "; ", F_wald$Res.Df[1],") = ", formatC(F_wald$F[2], digits = 2, format = "f"))`; $p<0.001$).

## Test for significance on each coefficient

Even by trying differe,t white's corrections, the results are not greatly different than when not accounting for Heteroskedasticity-consistent variance-covariance matrix

```{r}
lmtest::coeftest(x = model, vcov = sandwich::vcovHC(model, type = "HC4"))
```

\clearpage

## Breusch-Pagan Test for heteroskedstic errors

Test results indicate we have heteroskedastic errors in this model:

```{r}
lmtest::bptest(model, studentize = T)
```

## Durbin-Watson Test for autocorrelation

This model indicates no autocorrelation (DW = 1.994, $p = 0.26$).

```{r}
lmtest::dwtest(model)
```

