---
title: "Assignment 2"
subtitle: "Spatial Regression Models"
author: "Juan Carlos Villaseñor-Derbez"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,
                      fig.height = 4,
                      fig.width = 6,
                      warning = F,
                      message = F)
```

```{r, echo = F}
suppressPackageStartupMessages({
  library(spdep)
  library(maptools)
  library(sf)
  library(cowplot)
  library(tidyverse)
})
```


```{r}
riverside <- read_sf(dsn = "./shp", layer = "LPA_Pop_Char_bg") %>% 
  filter(countyname == "Riverside") %>% 
  mutate(location = case_when(LPAgrp == 0 ~ "None",
                              LPAgrp == 1 ~ "Rural",
                              LPAgrp == 2 ~ "Exurb",
                              LPAgrp == 3 ~ "Suburb",
                              TRUE ~ "Center"),
         LPAgrp = fct_relevel(location, "Center", "Suburb", "Exurb", "Rural"))
```

# Visualize the data

```{r, fig.height = 2, fig.width = 6}
ggplot(riverside) +
  geom_sf(aes(fill = drvalM), color = "transparent") +
  theme_minimal() +
  scale_y_continuous(breaks = c(33.5, 34)) +
  scale_fill_gradientn(colors = colorRamps::matlab.like(20))
```

\clearpage

# Fit the models

## Non spatial OLS

The model is specified by:

$$drvalM = \beta_0 + \sum_{s = 1}^7\sigma_sHHSIZE_s+\sum_{v = 1}^8\gamma_v HHVEH_v + \sum_{l = 1}^4\lambda_lLPAgrp_l +\epsilon$$

Which assumes that outcome variables are a function of household size, household vehicles, and location of the block group

```{r}
frmla <- formula("drvalM ~ HHSIZE1 + HHSIZE2 + HHSIZE3 + HHSIZE4 +
                 HHSIZE5 + HHSIZE6 + HHSIZE7 + HHVEH1 + HHVEH2 +
                 HHVEH3 + HHVEH4 + HHVEH5 + HHVEH6 + HHVEH7 + HHVEH8 + LPAgrp")

NonSpOLS <- lm(formula = frmla, data = riverside)
```

## Define 10 nearest neighbors for spatial models

```{r}
ids <- rownames(riverside)

knn <- riverside %>% 
  as("Spatial") %>% 
  coordinates() %>% 
  knearneigh(k=10) %>% 
  knn2nb(knn = ., row.names = ids) %>% 
  nb2listw(style = "W")
```

## Spatial Lag

The spatial lag model assumes that, in addition to variables in the ordinary least squares regression, the $Y$ values of a blockgroup are also influenced by the $Y$ values of their neighboring block groups.

$$drvalM = \beta_0 + \sum_{s = 1}^7\sigma_sHHSIZE_s+\sum_{v = 1}^8\gamma_v HHVEH_v + \sum_{l = 1}^4\lambda_lLPAgrp_l +\rho WY + \epsilon$$

```{r}
SpLag <- lagsarlm(formula = frmla, data = riverside, listw = knn)
```

## Spatial Error

The spatial error model assumes that the observations of $Y$ are correlated with the neighboring block groups' errors. That is, the $\epsilon$ term in the non-spatial regression model is correlated with $]epslion$ of the neighbors. Thismodel is specified as

$$drvalM = \beta_0 + \sum_{s = 1}^7\sigma_sHHSIZE_s+\sum_{v = 1}^8\gamma_v HHVEH_v + \sum_{l = 1}^4\lambda_lLPAgrp_l + \upsilon$$

Where $\upsilon$ is given by $\lambda W \upsilon + \epsilon$

These first three models can be examined in table 1.

```{r}
SpaErr <- errorsarlm(formula = frmla, data = riverside, listw = knn)
```

## SARAR

Finally, the SARAR model assumes that $Y$ is explained by a set of explanatory variables, neighboring $Y$s and neighboring errors:

$$drvalM = \beta_0 + \sum_{s = 1}^7\sigma_sHHSIZE_s+\sum_{v = 1}^8\gamma_v HHVEH_v + \sum_{l = 1}^4\lambda_lLPAgrp_l +\rho WY + \upsilon$$

Where $\upsilon$ is given by $\lambda W \upsilon + \epsilon$

```{r}
sarar <- sacsarlm(formula = frmla, data = riverside, listw = knn)
```

## Spatial Lag with Lagged explanatory variables

```{r}
SpLagLag <- lagsarlm(formula = frmla, data = riverside, listw = knn, type= "mixed")
```


## SARAR with Lagged explanatory variables

```{r}
sararlag <- sacsarlm(formula = frmla, data = riverside, listw = knn, type= "sacmixed")
```

# Model interpretations and findings

Overall model interpretation is very similar. While the `impacts` function provides a more clear interpretation of the effects by separating the direct from the indirect interactions, we can visualize the effects with a termplot (Figure 2). Full model coefficients and goodness of fit are provided in table 1 and the printed `impacts`.

We observe that groupblock where households are located further away, have higher values of `drvalM` (Fig 2A). The SARAR and SARAR lagged model have higer estimates for these coefficients. Household size has different effects on `drvalM` (Fig 2B). We observe that households made up from 1 and 2 people drive less. Then, households made of 3 and 4 people drive more, and then the estimate is again reduced. Increasing the number of vehicles in a household block group also resulted in increases in `drvalM`, with exception of households with 7 vehicles (Fig 2C).

\clearpage

# Comparing the models

```{r, fig.width = 5, fig.height = 8, echo = F}
get_terms <- function(model){
  tibble(term = names(coef(model)), estimate = coef(model)) %>% 
    filter(!term %in% c("rho", "lambda")) %>% 
    mutate(std.error = model$rest.se)
}

spatial <- list(c(SpLag, y = "SpLag"), c(SpaErr, y = "SpErr"), c(sarar, y = "SARAR"), c(SpLagLag, "SPLAGLAG"), c(sararlag, y = "SARARlag")) %>% 
  map_df(get_terms) %>% 
  mutate(model = c(rep("SpLag", 19), rep("SpErr", 19), rep("SARAR", 19), rep("SARARlaglag", 37), rep("SARARlag", 37)))

pd <- position_dodge(width = 0.5)

LPAgrp <- broom::tidy(NonSpOLS) %>% 
  select(term, estimate, std.error) %>% 
  mutate(model = "NonSpOLS") %>% 
  rbind(spatial) %>% 
  filter(grepl(pattern = "LPAgrp", x = term),
         !grepl(pattern = "lag", x = term)) %>% 
  ggplot(aes(x = term, y = estimate, color = model, fill = model)) +
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error), width = 0, size = 1, position = pd) +
  geom_point(size = 3, shape = 21, color = "black", position = pd) +
  coord_flip() +
  theme(legend.position = "none")

HHsize <- broom::tidy(NonSpOLS) %>% 
  select(term, estimate, std.error) %>% 
  mutate(model = "NonSpOLS") %>% 
  rbind(spatial) %>% 
  filter(grepl(pattern = "HHSIZE", x = term),
         !grepl(pattern = "lag", x = term)) %>% 
  ggplot(aes(x = term, y = estimate, color = model, fill = model)) +
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error), width = 0, size = 1, position = pd) +
  geom_point(size = 3, shape = 21, color = "black", position = pd) +
  coord_flip() +
  theme(legend.position = c(0,0.5))

HHveh <- broom::tidy(NonSpOLS) %>% 
  select(term, estimate, std.error) %>% 
  mutate(model = "NonSpOLS") %>% 
  rbind(spatial) %>% 
  filter(grepl(pattern = "HHVEH", x = term),
         !grepl(pattern = "lag", x = term)) %>% 
  ggplot(aes(x = term, y = estimate, color = model, fill = model)) +
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error), width = 0, size = 1, position = pd) +
  geom_point(size = 3, shape = 21, color = "black", position = pd) +
  coord_flip() +
  theme(legend.position = "none")

cowplot::plot_grid(LPAgrp, HHsize, HHveh, ncol = 1, labels = "AUTO")

```

```{r, results = "asis", echo = F}
stargazer::stargazer(NonSpOLS, SpLag, SpaErr,
                     single.row = T,
                     type = "latex", header = F, title = "Coefficient estimates for a non-spatial OLS regression, a regression with spatial lags, and a regression with spatial errors. Type of model is indicated above each column.")
```

\clearpage

```{r}
impacts(sarar, listw = knn)
impacts(SpLagLag, listw = knn)
impacts(sararlag, listw = knn)
```
