---
title: "Lab 1"
subtitle: "Intro to R and RStudio"
author: "Juan Carlos Villaseñor-Derbez"
date: "16 de enero de 2018"
output:
  html_document:
    code_folding: show
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages

```{r}
suppressPackageStartupMessages({
  library(psych)
  library(corrplot)
  library(tidyverse)
})
```

# Load data

```{r}
SmallHHfile <- read_csv("SmallHHfile.csv", col_types = cols())
```

# Describe data

## The basics

```{r}
# inspect the data we imported
# View(SmallHHfile)

# display the data.frame
str(SmallHHfile)

# Provide a summary of the descriptive statsitics of all the variables
summary(SmallHHfile)
```

## Describe the data with `psych`

```{r}
describe(SmallHHfile)
#
# package psych also computes descriptive statistics by groups of one or more variables
# not very pretty but does the job

describeBy(SmallHHfile$HTRIPS, SmallHHfile$DOW)
```

## Analysis that outputs a matrix of results and stores them in Mat1

```{r}
Mat1 <- describeBy(SmallHHfile$HTRIPS,list(SmallHHfile$DOW,SmallHHfile$HHVEH),
                   mat=TRUE,digits=2)  #matrix output

# display the matrix Mat1
Mat1
```

## With `skimr`

```{r}
skimr::skim(SmallHHfile)
```

# Visualize data

## `ggplot2`

```{r}
ggplot(SmallHHfile, aes(HHSIZ, HTRIPS, colour = INCOM)) + 
  geom_point()


m<-ggplot(SmallHHfile, aes(HTRIPS))

m + geom_histogram(binwidth = 1)

m + geom_histogram(aes(fill = ..count..), binwidth = 1) +
  scale_fill_gradient("Count", low = "green", high = "red")

```

## `ggplot_2` with `facet_*`

```{r}
m <- m + geom_histogram(binwidth = 1)
m + facet_grid(. ~ center)
```

## Color categories

```{r}
k <- ggplot(SmallHHfile, aes(factor(HTRIPS), fill = factor(HHVEH)))
k + geom_bar()
```

## Other plots

### Boxplots

```{r}
boxplot(SmallHHfile$HTRIPS)
```

#### Switching off the drawing of outliers

```{r}
boxplot(SmallHHfile$HHSIZ, outline = FALSE)
```

#### Show boxplots of household size for each group of household by different number of cars

```{r}
boxplot(SmallHHfile$HHSIZ ~ SmallHHfile$HHVEH, outline = TRUE)
```

#### Labeling the categories of household cars and the overall graph of boxplots

```{r}
# Note how names = c() is used to assign labels
boxplot(SmallHHfile$HHSIZ ~ SmallHHfile$HHVEH,
        outline = TRUE,
        main="Household Cars by Household Size",
        names = c("0car","1car","2car","3car","4car","5car","6car","7car","8car"))
```

### Histogram + boxplot in a single figure

```{r}
#
nf <- layout(mat = matrix(c(1,2),2,1, byrow=TRUE), height = c(3,1))

par(mar=c(3.1, 3.1, 1.1, 2.1))

hist(SmallHHfile$TotDist,
     col=rgb(0,1,0,0.9),
     breaks=400,
     xlim=c(0,800), 
     ylim=c(0,16000),
     xlab="Number of Travel Miles",
     main=" Household Vehicle Miles of Daily Travel")

boxplot(SmallHHfile$TotDist, horizontal=TRUE,  outline=TRUE,  ylim=c(0,800), frame=F, col = "green1")
box()
```

#### Same as above but no outliers for TotDist

```{r}
nf <- layout(mat = matrix(c(1,2),2,1, byrow=TRUE), height = c(3,1))

par(mar=c(3.1, 3.1, 1.1, 2.1))

hist(SmallHHfile$TotDist,
     col=rgb(0,1,0,0.9),
     breaks=400,
     xlim=c(0,800), 
     ylim=c(0,16000),
     xlab="Number of Travel Miles",
     main=" Household Vehicle Miles of Daily Travel")

boxplot(SmallHHfile$TotDist, horizontal=TRUE,  outline=FALSE,  ylim=c(0,800), frame=F, col = "green1")

box()
```

# Measures of association

## pairwise correlations among all variables round to 2 decimal places

```{r}
COVMAT = cov(SmallHHfile)
round(COVMAT, 3)
```

```{r}
names(SmallHHfile) # check the names of the columns
NEWFRAME <-select(SmallHHfile, INCOM:HTRIPS) # select the columns from INCOM to HTRIPS
corstats<-cor(NEWFRAME) # compute correlations of these few variables
round(corstats, 2) # print correlation with 2 decimal places
# pairwise covariances among all variables 
corrplot(corstats, method = "ellipse", type = "lower")
COVMAT = cov(NEWFRAME)
round(COVMAT, 3)
```


```{r}
MONDAY <-filter(SmallHHfile, Mon == 1)
NEWMON <-select(MONDAY, INCOM:HTRIPS)
corstats<-cor(NEWMON)
round(corstats, 2)
corrplot(corstats, method = "ellipse", type = "lower")

covstats<-cov(NEWMON)
round(covstats, 2)
```


